Toronto's Covid-19 Impact: Effect of Age and Gender on Outcomes

### 2024-09-24 paper/paper.qmd
updating csv files and writing the paper

(\scriptsize Table Separated using Tips from [@stack_tip] and LLMs)

@@

```{r}
#| label: tbl-clean
#| tbl-cap: "COVID-19 Case Cleaned Data Part 2"
#| echo: false
#| warning: false
#| message: false
#| 
# Cleaned Data Table (Table 2)
cleaned_covid_data %>%
  head(5) %>%
  kable(col.names = c("x_id", "age_group", "client_gender", "reported_date", "ever_in_icu", "ever_hospitalized", "ever_intubated", "outcome"),
        booktabs = TRUE)
```

@@

## Cases over Time

@fig-cumulativecases displays the progression of COVID-19 cases in Toronto from January 2020 to March 2024.

```{r}
#| label: fig-cumulativecases
#| fig-cap: "COVID-19 Case Progression Over Time, Toronto"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

# Step 3: Plot cumulative cases over time
ggplot(cleaned_covid_data, aes(x = reported_date, y = x_id)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "Cumulative COVID-19 Cases Over Time in Toronto",
       x = "Date",
       y = "COVID-19 Cases") +
  scale_x_date(limits = as.Date(c("2020-01-01", "2024-03-31")))

@@

```{r}
#| label: fig-treatmentoutcomes
#| fig-cap: "Level of Hospitalisation vs Mortality Rates"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

# Step 2: Plot the mortality rate as a bar graph
ggplot(hospitalised_mortality_data, aes(x = Condition, y = Mortality_Rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mortality Rate by Condition (Intubated, Hospitalized, ICU)",
       x = "Condition",
       y = "Mortality Rate (%)") +
  theme_minimal() +
  ylim(0, 100)  # Set y-axis limits to 0-100 for percentage

@@

```{r}
#| label: fig-agegroupmortality
#| fig-cap: "Age Group vs Mortality Rates"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

```{r}
#| label: fig-agegrouphospitalisation
#| fig-cap: "Chance of Hospitalisation by Age Group"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

# Step 4: Plot Age Group vs Chance of Hospitaliation, Intubation, or ICU as a grouped bar chart
ggplot(long_hosp_data, aes(x = Age_Group, y = Rate, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hospitalisation, Intubation, ICU Admission by Age Group",
       x = "Age Group",
       y = "Percentage (%)") +
  theme_minimal() +
  ylim(0, 25) +  # Set y-axis limits to 0-100%
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```
## Gender vs Outcomes

```{r}
#| label: fig-gendermortality
#| fig-cap: "COVID-19 Mortality by Gender"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

```
# Discussion {#sec-discussion}

@@

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

@@

@@

@@

Weaknesses and next steps should also be included.

@@

>>>>

(Table Separated using Tips from [@stack_tip] and LLMs)

@@

\scriptsize
```{r}
#| label: tbl-clean
#| tbl-cap: "COVID-19 Case Cleaned Data Part 2"
#| echo: false
#| warning: false
#| message: false
#| 
# Cleaned Data Table (Table 2)
cleaned_covid_data %>%
  head(5) %>%
  kable(col.names = c("x_id", "age_group", "client_gender", "reported_date", "ever_in_icu", "ever_hospitalized", "ever_intubated", "outcome"),
        booktabs = TRUE)
```
\normalsize

@@

## General Results

@fig-cumulativecases displays the progression of COVID-19 cases in Toronto from January 2020 to March 2024. @fig-treatmentoutcomes importantly highlights mortality rates by level of hospitalisation. That is, the mortality rates for those hospitalised, put in the ICU, and/or intubated.

```{r}
#| label: fig-cumulativecases
#| fig-cap: "COVID-19 Case Progression Over Time, Toronto"
#| fig.width: 5
#| fig.height: 2.75
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

# Step 3: Plot cumulative cases over time
ggplot(cleaned_covid_data, aes(x = reported_date, y = x_id)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "COVID-19 Cases Over Time, Toronto",
       x = "Date",
       y = "COVID-19 Cases") +
  theme(
    plot.title = element_text(size = 12),        # Reduce the title font size
    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
  ) +
  scale_x_date(limits = as.Date(c("2020-01-01", "2024-03-31")))

@@

```{r}
#| label: fig-treatmentoutcomes
#| fig-cap: "Level of Hospitalisation vs Mortality Rates"
#| fig.width: 5
#| fig.height: 2.75
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

# Step 2: Plot the mortality rate as a bar graph
ggplot(hospitalised_mortality_data, aes(x = Condition, y = Mortality_Rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Mortality by Condition (Intubated, Hospitalized, ICU)",
       x = "Condition",
       y = "Mortality Rate (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),        # Reduce the title font size
    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
  ) +
  ylim(0, 100)  # Set y-axis limits to 0-100 for percentage

@@

@fig-agegroupmortality shows the differences in mortality rates for different age groups at the time of contracting COVID-19. @fig-agegrouphospitalisation shows the hospitalisation, ICU, or intubation rates for different age groups. 
```{r}
#| label: fig-agegroupmortality
#| fig-cap: "Age Group vs Mortality Rates"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

```{r}
#| label: fig-agegrouphospitalisation
#| fig-cap: "Hospitalisation, Intubation, ICU Admission Rate by Age"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

# Step 4: Plot Age Group vs Chance of Hospitaliation, Intubation, or ICU as a grouped bar chart
ggplot(long_hosp_data, aes(x = Age_Group, y = Rate, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hospitalisation, Intubation, ICU Admission Rate by Age",
       x = "Age Group",
       y = "Percentage (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),        # Reduce the title font size
    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
    legend.title = element_text(size = 8),       # Reduce the legend title font size
    legend.text = element_text(size = 6)         # Reduce the legend text font size
  ) +
  ylim(0, 25) +  # Set y-axis limits to 0-100%
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```
\newpage

## Gender vs Outcomes

@fig-gendermortality graphs COVID-19 mortality rates by gender. @fig-genderhospitalisation graphs hospitalisation, ICU and/or incubation rates by gender. That is, how likely a person who has contracted COVID-19 is likely to be in any of these conditions, by gender.
```{r}
#| label: fig-gendermortality
#| fig-cap: "COVID-19 Mortality by Gender"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

```{r}
#| label: fig-genderhospitalisation
#| fig-cap: "Hospitalisation, ICU, Intubation Rate by Gender"
#| fig.width: 5
#| fig.height: 3
#| fig.align: center
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"
# Step 1: Calculate the rates for hospitalization, intubation, and ICU admission by gender
genders <- c("MALE", "FEMALE", "OTHER")

hospitalization_data_gender <- data.frame(
  Client_Gender = genders,
  Hospitalization_Rate = sapply(genders, function(gender) {
    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
    calc_condition_rate(df_gender, "ever_hospitalized")
  }),
  Intubation_Rate = sapply(genders, function(gender) {
    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
    calc_condition_rate(df_gender, "ever_intubated")
  }),
  ICU_Rate = sapply(genders, function(gender) {
    df_gender <- cleaned_covid_data %>% filter(client_gender == gender)
    calc_condition_rate(df_gender, "ever_in_icu")
  })
)

# Step 2: Reshape the data to long format for grouped bar plot
long_data_gender <- hospitalization_data_gender %>%
  pivot_longer(cols = c(Hospitalization_Rate, Intubation_Rate, ICU_Rate),
               names_to = "Condition",
               values_to = "Rate")

# Step 3: Plot Gender vs Chance of Hospitalization, Intubation, or ICU as a grouped bar chart
ggplot(long_data_gender, aes(x = Client_Gender, y = Rate, fill = Condition)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Hospitalization, Intubation, or ICU by Gender",
       x = "Gender",
       y = "Percentage (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12),        # Reduce the title font size
    axis.title.x = element_text(size = 10),      # Reduce the x-axis title font size
    axis.title.y = element_text(size = 10),      # Reduce the y-axis title font size
    axis.text.x = element_text(size = 8),        # Reduce the x-axis text (labels) font size
    axis.text.y = element_text(size = 8),        # Reduce the y-axis text (labels) font size
    legend.title = element_text(size = 8),       # Reduce the legend title font size
    legend.text = element_text(size = 6)         # Reduce the legend text font size
  ) +
  ylim(0, 10) +  # Set y-axis limits to 0-100%
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))  # Keep x-axis labels horizontal
```
# Discussion {#sec-discussion}

@@

Understanding the effect age has on COVID-19 is extremely important both for potential victims and those in charge of policy to put in place correct and appropriate health measures for different ages.

Looking at @fig-agegroupmortality, we see a clear message - age group and mortality are positively correlated. More specifically, mortality past the ages of 50-59 shoots from under half a percent to 1.72% for 60-69 years of age. This only worsens as individuals get older. For the age groups 70-79 and 80-89, the mortality rate if one contracts COVID-19 shoots to 4.14% and 7.50% respectively. Moreover, those who contract COVID-19 at ages of 90 and above have a mortality rate of almost 10% (9.62%).

In contrast, those younger than 50 at the time of contraction have mortality rates under 0.5%. In fact, those 19 and younger have a mortality rate of only 0.007%. Clearly, age and mortalty are very strongly correlated.

Perhaps of more relevance are chances of hospitalisation, admission to the ICU, and intubation. In reference to @fig-agegrouphospitalisation, we see that the older population is much more susceptible to having more severe cases of COVID-19. For all ages above 70, the chance of hospitalisation is above 15%, peaking at 19.46% for 80-89 year olds. Compared to the younger population, this is a stark difference. Those under 30 have hospitalisation rates of under 30%. This increases with age: hospital admission rates for those with COVID-19 is 1.15% for 30-39 year olds, 2.13% for 40-49 year olds, 4.15% for 50-59 year olds, and 8.91% for 60-69 year olds.

discussion about weaker immune systems with age (bring in some literature review)

@@

Similar to age, understanding differences in outcomes for genders matter greatly to virologists and as a matter of public safety. Looking at @fig-gendermortality, we find that males that contract COVID-19 have a much higher mortality rate when compared to females (1.47% vs 1.06%). According to this data, males who contract COVID-19 are almost 1.4 times more likely to die than females.

Similarly, looking at @fig-genderhospitalisation, we see that males are more likely to be hospitalised, put in the ICU, and intubated.

discussion about how Covid-19 affects the male body vs female; lit and scientific review; is this a worldwide trend?

@@

Reference the importance of @fig-treatmentoutcomes - those intubated have almost a 50% chance of death. (what?!)

@@

Measurement weaknesses (stopped collecting data), might be important to look at selection biases (are there more male cases than females? what about Other genders? are they affect results?

@@

ChatGPT Data Analyst was used to generate code and help fix bugs for this assignment. A full LLM Disclosure can be found on the GitHub Repository under "Other - LLM - usage.txt"
